What do they see? Testing perceptual accuracy in a U.S. general population survey using stacked bar charts

### 2023-10-27 SDSS-2023/testing-charts-accuracy.tex
modified tex file, rendered to pdf

\begin{document}
\maketitle
\begin{abstract}
The use of visuals is a key component in scientific communication.
Decisions about the design of a data visualization should be informed by
what design elements best support the audience's ability to perceive and
understand the components of the data visualization. We build on the
foundations of Cleveland and McGill's work in graphical perception,
employing a large, nationally-representative, probability-based panel of
survey respondents to test perception in stacked bar charts. Our
findings provide actionable guidance for data visualization
practitioners to employ in their work.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, boxrule=0pt, sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Cleveland and McGill's work -- while a foundational user study in
graphical perception -- utilized a small convenience sample, consisting
of only a few individuals recruited from among the authors' coworkers
and their spouses. Heer and Bostock (2010) reproduced Cleveland and
McGill's rankings using a larger sample from a crowd sourcing platform,
employing a total of 82 Amazon Mechanical Turk (mTurk) workers for the
study. Crowd sourced samples were shown by Borgo et al. (2017) to be
biased towards more male, younger, and relatively well-educated
individuals relative to the adult U.S. population as a whole; generally,
convenience and opt-in samples are not representative of the general
population. Further, higher education individuals are likely to have
prior exposure to data visualization or data topics in higher education,
which may not be the case for other groups within the population;
existing differences in exposure to data may drive differences in
results across groups. The general public is a common target audience
for data visualization and scientific communication work; results of
polls, surveys, and scientific work are presented often in the news,
shared on social media, or posted online or in print materials. It is
imperative that data visualizations used to communicate these topics are
designed with the goal of reaching the entire population. To that end,
an inclusive, representative approach to studying how viewers perceive
information presented in charts is critical to understand how best to
communicate with the general public.

In this work, we seek to answer whether it is possible to reproduce some
of the previous findings in the context of a survey with a large,
nationally-representative set of respondents. Specifically, we present
viewers with structural variations of bar charts and ask them to answer
questions comparing the size of elements within those charts. We employ
a probability-based survey panel and run a series of perception tests
with nationally-representative samples of respondents from that panel.
The advantage of using a probability-based approach is two-fold. First,
we have access to a large sample of survey participants and thus have
greater power in making inference about graphical perceptional
abilities. Second, the sample is representative of the general adult
public in the U.S., and this allows us to test whether prior results
from convenience samples hold with a nationally representative sample
and whether there are differences in those results across demographic
subgroups. Within that context we focus on the following research
questions:

@@

\autoref{fig-alignment}(a) displays all responses along the binary
correctness measure, separated by whether the stimulus was an aligned
task or unaligned task. We can see that levels of accuracy for all
responses are significantly higher for the `easier' (aligned) task, with
about twice as many respondents correctly selecting the larger of the
two marked elements.

@@

{\centering \includegraphics[width=0.9\textwidth,height=\textheight]{./figures/fig-alignment-1.png}

@@

\caption{\label{fig-alignment}On the left (a), a stacked barchart shows
the number of respondents with correct (green) and incorrect (grey)
responses to the two comparison questions. When tiles are aligned along
the same axis, more than twice the number of responses are correct. The
shaded area along the top of the green bars corresponds to 95\%
confidence intervals around (marginal) correct responses. On the right
(b), a parallel coordinate plot shows all combinations of responses.
There's a huge asymmetry in the number of responses where participants
answered only one of the questions correctly.}

@@

\hypertarget{tbl-zoom-123}{}
\begin{longtable}{lrrrr}
\caption{\label{tbl-zoom-123}Coefficients for logistic regression of zooming by task }\tabularnewline

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
The use of visuals is a key component in scientific communication.
Decisions about the design of a data visualization should be informed by
what design elements best support the audience's ability to perceive and
understand the components of the data visualization. We build on the
foundations of Cleveland and McGill's work in graphical perception,
employing a large, nationally-representative, probability-based panel of
survey respondents to test perception in stacked bar charts. Our
findings provide actionable guidance for data visualization
practitioners to employ in their work.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

Cleveland and McGill's work -- while a foundational user study in
graphical perception -- utilized a small convenience sample, consisting
of only a few individuals recruited from among the authors' coworkers
and their spouses. Heer and Bostock (2010) reproduced Cleveland and
McGill's rankings using a larger sample from a crowd sourcing platform,
employing a total of 82 Amazon Mechanical Turk (mTurk) workers for the
study. Crowd sourced samples were shown by Borgo et al. (2017) to be
biased towards more male, younger, and relatively well-educated
individuals compared to the general adult U.S. population. Presumably,
individuals with a higher education are more likely to have prior
exposure to data visualization and might be more familiar with
quantitative statements than less well-educated groups within the
population; other pre-existing differences between population groups may
also drive differences in understanding or interpreting charts. This
raises the question of whether conclusions drawn from these studies hold
for the general population -- a common target audience for data
visualization and scientific communication work. Results of polls,
surveys, and scientific work are presented visually in the news, shared
on social media, posted online or in print materials. It is imperative
that data visualizations used to communicate quantitative information
are designed with the goal of reaching the entire population. To that
end, an inclusive, representative approach to studying how viewers
perceive information presented in charts is critical to understand how
best to communicate with the general public.

In this work, we seek to answer whether it is possible to reproduce some
of the previous findings in the context of a survey with a large,
nationally-representative set of respondents. Specifically, we present
viewers with structural variations of bar charts and ask them to answer
questions comparing the size of elements within those charts. We employ
a probability-based survey panel and run a series of perception tests
with nationally-representative samples of respondents from that panel.
The advantage of using a probability-based approach is two-fold: First,
the relatively larger sample of survey participants provides greater
statistical power in drawing inference about graphical perceptional
abilities. Second, the sample is representative of the general adult
public in the U.S., permitting an assessment inhowfar prior results from
convenience samples hold with a nationally representative sample and
whether there are differences in those results across demographic
subgroups. Within that context we focus on the following research
questions:

@@

\autoref{fig-alignment} displays all responses along the binary
correctness measure, separated by whether the stimulus was an aligned
task or unaligned task. We can see that levels of accuracy for all
responses are significantly higher for the `easier' (aligned) task, with
about twice as many respondents correctly selecting the larger of the
two marked elements.

@@

{\centering \includegraphics[width=0.4\textwidth,height=\textheight]{./figures/fig-alignment-1.png}

@@

\caption{\label{fig-alignment}Participants' response pattern to both
questions. The coloring reflects the number of correct responses (two in
full green, zero in grey, and one correct answer in grey-green). When
tiles are aligned along the same axis, more than twice the number of
respondents correctly identifies the larger tile. There's a huge
asymmetry in the number of responses where participants answered only
one of the questions correctly.}

@@

%\hypertarget{tbl-zoom-123}{}
\begin{longtable}{lrrrr}
\caption{\label{tbl-zoom-123}Coefficients for logistic regression of zooming by task }\tabularnewline

@@

\hypertarget{supplementary-material}{%
\section{Supplementary Material}\label{supplementary-material}}

The source code and data used in this paper are available on a GitHub
repository at \url{https://github.com/kiegan/testing-charts-jds}. In
addition, we provide an example stimulus image in that repository.
